A fast and Low-Overhead Technique to Secure Programs Against Integer Overflows

### 2012-09-09 doc/Manuscript/CGO/paper.tex
Polished the paper a little bit.

% TODO: explain that we have not observed overflows in the safe instructions.

@@

\begin{abstract}
The integer primitive type has upper and lower bounds in many programming
languages, including C, C++ and Java.
This limitation might lead programs that manipulate large integer numbers to
produce unexpected results due to overflows.
There exists a plethora of works that instrument programs to track the occurrence
of these overflows.
In this paper we present an algorithm that relies on static range analysis to
eliminate this instrumentation whenever possible.
We present a novel live range splitting strategy that refine the precision of
our range analysis.
We have used this algorithm to improve the quality of a dynamic instrumentation
library that we have implemented in LLVM.
This framework has been used to detect overflows in hundreds of C/C++ programs.
As a testimony of its effectiveness, our range analysis has been able to eliminate
19\% of all the overflow checks necessary to secure the C programs in the
SPEC CPU 2006 benchmark suite.
\end{abstract}

@@

% What are integer overflows.
The most popular programming languages, including C, C++ and Java, limit the
size of primitive numeric types.
For instance, the \texttt{int} type, in C++, ranges from $-2^{31}$ to
$2^{31}-1$.
Consequently, there exist numbers that cannot be represented by these types.
In general, these programming languages resort to a {\em wrapping-arithmetics}
semantics~\cite{Warren02} to perform integer operations.
If a number $n$ is too large to fit into a primitive data type $T$, then $n$'s
value wraps around, and $n \mbox{module} T_{max}$ wounds up represented
instead.
There are situations in which this semantics is acceptable~\cite{Dietz12}.
For instance, programers might rely on this behavior to implement hash functions
and random number generators.
On the other hand, there exist also situations in which this behavior might
lead a program to produce unexpected results.
As an example, in 1996, the Ariane 5 rocket was lost due to an arithmetic
overflow -- a bug that resulted in a loss of more than US\$370 million~\cite{Dowson97}.

@@

% Our contribution: range analysis.
This paper describes the range analysis algorithm that
we have developed to eliminate overflow checks in instrumented programs.
As we show in Section~\ref{sec:range}, our algorithm has three core insights.
Firstly, in Section~\ref{sub:micro} we propose a three-phases approach to extract
information from comparisons between variables, e.g., $x < y$.
Previous algorithms either deal with these comparisons via
expensive relational analyses~\cite{Cousot78,Lakhdar11,Mine06}, or only
consider comparisons between variables and
constants~\cite{Mahlke01,Patterson95,Stephenson00}.
Secondly, in Section~\ref{sub:prop} we show how we rely on strongly connected
components to achieve speed and precision.
It is well-known that this technique is effective in speeding up constraint
resolution~\cite[Sec 6.3]{Nielson99}.
However, given our three-phases approach, we improve precision
by solving strong components in topological order.
Finally, in Section~\ref{sub:splitting} we propose a new program representation that
is more precise than other intermediate forms used to solve range
analysis sparsely.
This new live range splitting strategy is only valid if the instrumented
program terminates whenever an integer overflow is detected.
If we cannot rely on this guarantee, then our more conservative live range splitting
strategy produces the program representation that Bodik
{\em et al.}~\cite{Bodik00} call Extended Static Single Assignment
form.
In Section~\ref{sec:exp} we show that an interprocedural implementation of our
algorithm analyzes programs with half-a-million lines of code in less than ten
seconds.
Furthermore, the speed and memory consumption of our range analysis grows linearly
with the program size.

@@

From the lattice $\varset{Z}$ we define the product lattice
$\varset{Z}^2$, which is defined as follows:
%
\begin{equation*}
\varset{Z}^2 = \{ \emptyset \} \cup \{[z_1, z_2] | \ z_1,z_2 \in \varset{Z},
\ z_1 \leq z_2, \  -\infty < z_2 \}
\end{equation*}
%
This interval lattice is partially ordered by the subset relation, which we
denote by ``$\sqsubseteq$".
Range intersection, ``$\sqcap$", is defined by:
\[
[a_1, a_2] \sqcap [b_1, b_2] =
\begin{cases}
[\mbox{max}(a_1, b_1), \mbox{min}(a_2, b_2)], \ \mbox{if} \ a_1 \leq b_1 \leq a_2  \\ \mbox{or} \ b_1 \leq a_1 \leq b_2 \\
[a_1, a_2] \sqcap [b_1, b_2] = \emptyset, \ \mbox{otherwise}
\end{cases}
\]
And range union, ``$\sqcup$", is given by:
\[
[a_1, a_2] \sqcup [b_1, b_2] = [\mbox{min}(a_1, b_1), \mbox{max}(a_2, b_2)]
\]

@@

\noindent
\textbf{Building the Constraint Graph. }
Given a set $\varset{C}$ of constraints, which define and/or use constraint
variables from a set $\varset{V}$, we build a constraint graph
$G = (\varset{C} \cup \varset{V}, E)$.
The vertices in \varset{C} are the {\em constraint nodes}, and the
nodes in \varset{V} are the {\em variable nodes}.
If $V \in \varset{V}$ is used in constraint $C \in \varset{C}$, then
we create an edge $\overrightarrow{VC}$.
If constraint $C \in \varset{C}$ defines variable $\varset{V} \in V$, then we
create an edge from $\overrightarrow{CV}$.
Figure~\ref{fig:ex_graph} shows the constraint graph that we build for the
program in Figure~\ref{fig:ex_eSSA_cgo}(b).
If $V$ is used by $C$ as the input of a future, then the edge from
$V$ to $C$ represents what Ferrante {\em et al.} call a {\em control
dependence}~\cite[p.323]{Ferrante87}.
We use dashed lines to represent these edges.
All the other edges denote {\em data dependences}~\cite[p.322]{Ferrante87}.
As we will show later, control dependence edges increase the precision of our
algorithm to solve future bounds.

@@

\noindent
\textbf{Propagating Ranges in Topological Order. }
After building the constraint graph, we find its strongly connected components.
We collapse these components in super nodes, and then proceed to propagate
ranges along the resulting digraph.
This approach is essential for scalability, because all the complexity of our
algorithm lies in the resolution of strong components.
Our tests show that the vast majority of the strongly connected components are
singletons.
For instance, 99.11\% of the SCCs in SPEC CPU 2006 gcc (\texttt{403.gcc}) have
only one node.
Moreover, the composite components usually contain a small number of nodes.
As an example, the largest component in \texttt{403.gcc}, has 2,131 nodes,
even though gcc's constraint graph has over one million nodes.
This large SCC exists due to a long chain of mutually recursive function calls.

@@

\noindent
\textbf{Future resolution: }
The next phase of the algorithm to determine intervals inside a strong
component consists in replacing futures by actual bounds, a task that we
accomplish by using the rules below:
%
\begin{eqnarray*}
\begin{array}{c}
\inferrule{Y = X \sqcap [l, \fun{ft}(V) + c] \\ \ub{I[V]} = u}
{Y = X \sqcap [l, u + c]} \mbox{\hspace{0.3cm}} u, c \in \mathbb{Z} \cup \{-\infty, \infty\}
\\
\\
\inferrule{Y = X \sqcap [\fun{ft}(V) + c, u] \\ \lb{I[V]} = l}
{Y = X \sqcap [l + c, u]} \mbox{\hspace{0.3cm}} l, c \in \mathbb{Z} \cup \{-\infty, \infty\}
\end{array}
\end{eqnarray*}
%
In order to correctly replace a future $\fun{ft}(V)$ that limits a constraint
variable $V'$, we need to have already applied the growth analysis onto $V$.
Had we considered only data dependence edges, then it would be possible
that $V'$ be analyzed before $V$.
However, because of control dependence edges, this case cannot happen.
The control dependence edges ensure that any topological ordering of the
constraint graph either places $V$ before $V'$, or places these nodes
in the same strongly connected component.
For instance, in Figure~\ref{fig:ex_graph}, variables $j_1$ and $i_t$ are in
the same SCC only because of the control dependence edges.
Figure~\ref{fig:ex_partition_grow_crop}(c) shows the result of resolving
futures in our running example.
The information that we acquire from the growth analysis is essential in this
phase.
For instance, the growth analysis has found out that the value stored in variable
\texttt{i}$_i$ can only increase.
Given that this variable is assigned the initial value zero, we can replace
$\fun{ft}(I_1)$ with this value.

\noindent
\textbf{Narrowing: } the last step that we apply on the strongly connected
component is the narrowing phase.
In this step we use values extracted from conditional tests to restrict the
bounds of the constraint variables.
We use the narrowing operator firstly proposed by Cousot and
Cousot~\cite{Cousot77}, which we show below:
%
\begin{equation*}
I[Y] =
\begin{cases}
\mbox{if} \ \lb{I[Y]} = -\infty  \ \mbox{and} \ \lb{e(Y)} > -\infty \ \mbox{then} \ [\lb{e(Y)}, \ub{I[Y]}] \\
\mbox{elif} \ \ub{I[Y]} = \infty \ \mbox{and} \ \ub{e(Y)} < \infty \ \mbox{then}
\ [\lb{I[Y]}, \ub{e(Y)}] \\
\mbox{elif} \ \lb{I[Y]} > \lb{e(Y)} \ \mbox{then} \ [\lb{e(Y)}, \ub{I[Y]}] \\
\mbox{elif} \ \ub{I[Y]} < \ub{e(Y)} \ \mbox{then} \ [\lb{I[Y]}, \ub{e(Y)}]
\end{cases}
\end{equation*}
%
Figure~\ref{fig:ex_partition_grow_crop}(d) show the result of our narrowing
operator in our running example.
Ranges initially improve due to the two conditional tests in the program.
Firstly, we have that $I[I_t] = I[I_1] \sqcap [-\infty, 98]$, which gives us
that $I[I_t] = [0, 98]$.
We also have that $I[J_t] = I[J_1] \sqcap [0, \infty]$, giving
$I[J_t] = [0, 99]$.
From these new intervals, we can narrow the ranges bound to the other constraint
variables.

@@

Figure~\ref{fig:ex_ir} compares the u-SSA form with the SSA and e-SSA
intermediate program representations.
We use the notation $v = \bullet$ to denote a definition of variable $v$, and
$\bullet = v$ to denote a use of that variable.
Figure~\ref{fig:ex_ir}(b) shows the example program converted to the SSA format.
The different definitions of variable $u$ have been renamed, and a
$\phi$-function re-combines these definitions into a single name.
The SSA form sparsifies a dataflow analysis that only extracts information from
the definition sites of variables, such as constant propagation.
Figure~\ref{fig:ex_ir}(c) shows the same program in e-SSA form.
This time we have renamed variable $v$ right after the conditional test where
this variable is used.
The e-SSA form serves dataflow analyses that acquire information from definition
sites and conditional tests.
Examples of these analyses include array bounds checking
elimination~\cite{Bodik00} and traditional implementations of range
analyses~\cite{Gough94,Patterson95}.
Finally, Figure~\ref{fig:ex_ir}(d) shows our example in u-SSA form.
The live range of variable $v_1$ has been divided right after its use.
This representation is useful to analyses that learn information from the way
that variables are used, and propagate this information forwardly.

@@

%SUB instrumentation
The instrumentation that we insert is mostly straightforward.
We discuss in the rest of this section a few interesting cases.
If we are dealing with an unsigned \textsc{Sub} instruction, e.g,
$x = o_1 \ -_u \ o_2$, then a single check is enough to detect the bad
behavior: $o_1 < o_2$.
This is sound because if $o_2$ is greater than $o_1$, then we assume that it is
a bug to try to represent a negative number in unsigned arithmetics.
Regarding multiplication, e.g., $x = o_1 \times o_2$, if $o_1 = 0$, then this
operation can never cause an overflow.
This test is necessary, because we check integer overflows in
multiplication via the inverse operation, e.g., integer division.
Thus, the test prevents a division by zero from happening.
The \textsc{Trunc} instruction, e.g., $x = \ \downarrow_n \ o_1$ assigns to $x$
the $n$ least significant bits of $o_1$.
The dynamic check, in this case, consists in expanding $x$ to
the datatype of $o_1$ and comparing the expanded value with $o_1$.
The LLVM IR provides instructions to perform these type expansions.
Note that our instrumentation catches any truncation that might result in
data loss, even if this loss is benign.
To make the dynamic checks more liberal, we give users the possibility of
disabling tests over truncations.

\paragraph{Practical Considerations: }
%Overview about the instrumentation technique
Our instrumentation library works by adding new instructions to
the target program.
Although the dynamic check depends on the instruction that is instrumented,
the general modus operandi is the same.
The dynamic tests check for overflows after they happen.
The code that we insert to detect the overflow diverts the program flow in case
such an event takes place.
Figure~\ref{fig:instrumented_cfg} shows an actual control flow graph, before
and after the instrumentation.
From this figure, it is clear that the instrumented program will be larger
than the original code.
Figure~\ref{fig:num_instructions} shows how many LLVM instructions are
necessary to instrument each arithmetic operation.
These numbers do not include the instructions necessary to handle the overflow
itself, e.g., block \texttt{\%11} in Figure~\ref{fig:instrumented_cfg}, as this
code is not in the program's main path.
Nevertheless, as we show empirically, this growth is not too large when
compared to the total size of our benchmarks, because most of the instructions
in these programs do not demand instrumentation.
Furthermore, none of the instructions used to instrument integer arithmetics
access memory.
Therefore, the overall slowdown that the instrumentation causes is usually
small.

@@

Which actions are performed once the overflow is detected depends on the
user, who has the option to overwrite the \texttt{handle\_ overflow}
function in Figure~\ref{fig:instrumented_cfg}.
Our library gives the user three options to handle them. 
First option: no-op.
This option allows us to verify the slowdown produced by the new instructions.
Second option: logging.
This is the standard option, and it preserves the behavior of the instrumented
program.
Whenever an overflow is detected, we print \texttt{Overflow detected in
FileName.cpp, line X.} in the standard error stream.
Third option: abort.
This option terminates the program once an overflow is detected.
Thus, it disallows undefined behavior due to integer overflows.

@@

\noindent
\textbf{Time and Memory Complexity: }
Figure~\ref{fig:TimeCorr} compares the time to run our range analysis with the size of
the input programs.
We show data for the 100 largest benchmarks in our test suite, in number
of variable nodes in the constraint graph.
We perform function inlining before running our analysis.
Each point in the X line corresponds to a benchmark.
We analyze the smallest benchmark in this set, \texttt{Prolangs-C/deriv2}, which
has 1,131 variable nodes in the constraint graph, in 20ms.
We take 15.91 sec to analyze our largest benchmark, \texttt{403.gcc}, which,
after function inlining, has 1,266,273 assembly instructions, and a
constraint graph with 679,652 variable nodes.
For this data set, the coefficient of determination $(R^2)$ is 0.967, which
provides very strong evidence about the linear asymptotic complexity of our
implementation.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/TimeCorr}
\end{center}
\caption{\label{fig:TimeCorr}
Correlation between program size (number of var nodes in constraint
graphs after inlining) and analysis runtime (ms).
Each point represents a benchmark.
Coefficient of determination = 0.967.
}
\end{figure}

The experiments also reveal that the memory consumption of our implementation
is linear with the program size.
Figure~\ref{fig:MemCorr} plots these two quantities together.
The linear correlation, in this case, is even stronger than that found in
Figure~\ref{fig:TimeCorr}, which compares runtime and program size: the
coefficient of determination is 0.9947.
The figure only shows our 100 largest benchmarks.
Again, SPEC \texttt{403.gcc} is the heaviest benchmark, requiring
265,588KB to run.
Memory includes stack, heap and the executable program code.

@@

\noindent
\textbf{Precision: }
Our implementation of range analysis offers a good tradeoff between
precision and runtime.
Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance,
takes about 25 minutes to go over a program with almost 900 basic blocks.
We analyze programs of similar size in less than one second.
We do not claim our approach is as precise as such algorithms, even though we
are able to find exact bounds to 4/5 of the examples presented
in~\cite{Lakhdar11}.
On the contrary, we present a compromise between precision and speed
that scales to very large programs.
Nevertheless, our results are far from being trivial.
We have implemented a dynamic profiler that measures, for each variable,
its upper and lower limits, given an execution of the target program.
Figure~\ref{fig:precision} compares our results with those measured
dynamically for the Stanford benchmark, which is publicly
available in the LLVM test suite.
We chose Stanford because these benchmarks do not read data from external
files; hence, imprecisions are due to library functions that we cannot
analyze.

@@

We have classified the bounds estimated by the static analysis into four
categories.
The first category, called $1$, contains tight bounds:
during program execution, the variable has been assigned an upper,
or lower limit, that equals the limit inferred statically.
The second category, called $n$, contains the bounds that are
within twice the value inferred statically.
For instance, if the range analysis estimates that a variable $v$ is in the
range $[0, 100]$, and during the execution the dynamic profiler finds that
its maximum value is $51$, then $v$ falls into this category.
The third category, $n^2$, contains variables whose actual value is within
a quadratic factor from the estimated value.
In our example, $v$'s upper bound would have to be at most $10$ for it to
be in this category.
Finally, the fourth category contains variables whose estimated value lays
outside a quadratic factor of the actual value.
We call this category {\em imprecise}, and it contains mostly the limits that
our static analysis has marked as either $+\infty$ or $-\infty$.
As we see in Figure~\ref{fig:precision}, 54.11\% of the lower limits that
we have estimated statically are exact.
Similarly, 51.99\% of our upper bounds are also tight.
The figure also shows that, on average, 37.39\% of our lower limits are
imprecise, and 35.40\% of our upper limits are imprecise.
This result is on pair with those obtained by more costly analysis, such as
Stephenson {\em et al.}'s~\cite{Stephenson00}.

>>>>

% TODO: explain that we have not observed overflows in the safe instructions.
% TODO: explain that u-SSA does not yield noticeable performance improvements.

@@

\begin{abstract}
The integer primitive type has upper and lower bounds in many programming
languages, including C, and Java.
These limits might lead programs that manipulate large integer numbers to
produce unexpected results due to overflows.
There exists a plethora of works that instrument programs to track the occurrence
of these overflows.
In this paper we present the first algorithm that uses static range analysis to
avoid this instrumentation whenever possible.
Our range analysis contains novel techniques, such as a live range splitting
strategy that refines the precision of our results.
We have used this algorithm to avoid some checks created by a dynamic instrumentation
library that we have implemented in LLVM.
This framework has been used to detect overflows in hundreds of C/C++ programs.
As a testimony of its effectiveness, our range analysis has been able to avoid
25\% of all the overflow checks necessary to secure the C programs in the
LLVM test suite.
This reduction has reduced 50\% of the runtime overhead of instrumentation.
\end{abstract}

@@

% What are integer overflows.
The most popular programming languages, including C, C++ and Java, limit the
size of primitive numeric types.
For instance, the \texttt{int} type, in C++, ranges from $-2^{31}$ to
$2^{31}-1$.
Consequently, there exist numbers that cannot be represented by these types.
In general, these programming languages resort to a {\em wrapping-arithmetics}
semantics~\cite{Warren02} to perform integer operations.
If a number $n$ is too large to fit into a primitive data type $T$, then $n$'s
value wraps around, and we obtain $n \ \mbox{module} \ T_{max}$.
There are situations in which this semantics is acceptable~\cite{Dietz12}.
For instance, programers might rely on this behavior to implement hash functions
and random number generators.
On the other hand, there exist also situations in which this behavior might
lead a program to produce unexpected results.
As an example, in 1996, the Ariane 5 rocket was lost due to an arithmetic
overflow -- a bug that resulted in a loss of more than US\$370 million~\cite{Dowson97}.

@@

% Our contribution: range analysis.
This paper describes the range analysis algorithm that
we have developed to eliminate overflow checks in instrumented programs.
As we show in Section~\ref{sec:range}, our algorithm has three core insights.
Firstly, in Section~\ref{sub:micro} we propose a three-phases approach to extract
information from comparisons between variables, e.g., $x < y$.
Previous algorithms either deal with these comparisons via
expensive relational analyses~\cite{Cousot78,Lakhdar11,Mine06}, or only
consider comparisons between variables and
constants~\cite{Mahlke01,Patterson95,Stephenson00}.
Secondly, in Section~\ref{sub:prop} we show how we rely on strongly connected
components to achieve speed and precision.
It is well-known that this technique is effective in speeding up constraint
resolution~\cite[Sec 6.3]{Nielson99}.
However, given our three-phases approach, we improve precision
by solving strong components in topological order.
Finally, in Section~\ref{sub:splitting} we propose a new program representation that
is more precise than other intermediate forms used to solve range
analysis sparsely.
This new live range splitting strategy is only valid if the instrumented
program terminates whenever an integer overflow is detected.
If we cannot rely on this guarantee, then our more conservative live range splitting
strategy produces the program representation that Bodik
{\em et al.}~\cite{Bodik00} call Extended Static Single Assignment
form.
In Section~\ref{sub:range} we show that an interprocedural implementation of our
algorithm analyzes programs with half-a-million lines of code in less than ten
seconds.
Furthermore, the speed and memory consumption of our range analysis grows linearly
with the program size.

@@

From the lattice $\varset{Z}$ we define the product lattice
$\varset{Z}^2$ as follows:
%
\begin{equation*}
\varset{Z}^2 = \{ \emptyset \} \cup \{[z_1, z_2] | \ z_1,z_2 \in \varset{Z},
\ z_1 \leq z_2, \  -\infty < z_2 \}
\end{equation*}
%
This interval lattice is partially ordered by the subset relation, which we
denote by ``$\sqsubseteq$".
Range intersection, ``$\sqcap$", is defined by:
\[
[a_1, a_2] \sqcap [b_1, b_2] =
\begin{cases}
[\mbox{max}(a_1, b_1), \mbox{min}(a_2, b_2)], \ \mbox{if} \ a_1 \leq b_1 \leq a_2  \\ \ \ \ \ \mbox{or} \ b_1 \leq a_1 \leq b_2 \\
[a_1, a_2] \sqcap [b_1, b_2] = \emptyset, \ \mbox{otherwise}
\end{cases}
\]
And range union, ``$\sqcup$", is given by:
\[
[a_1, a_2] \sqcup [b_1, b_2] = [\mbox{min}(a_1, b_1), \mbox{max}(a_2, b_2)]
\]

@@

\noindent
\textbf{Building the Constraint Graph. }
Given a set $\varset{C}$ of constraints, which define and/or use constraint
variables from a set $\varset{V}$, we build a constraint graph
$G = (\varset{C} \cup \varset{V}, E)$.
The vertices in \varset{C} are the {\em constraint nodes}, and the
nodes in \varset{V} are the {\em variable nodes}.
If $V \in \varset{V}$ is used in constraint $C \in \varset{C}$, then
we create an edge $\overrightarrow{VC}$.
If constraint $C \in \varset{C}$ defines variable $\varset{V} \in V$, then we
create an edge $\overrightarrow{CV}$.
Figure~\ref{fig:ex_graph} shows the constraint graph that we build for the
program in Figure~\ref{fig:ex_eSSA_cgo}(b).
Our algorithm introduces the notion of {\em future ranges}, which we use to extract
range information from comparisons between variables.
In Section~\ref{sub:splitting} we explain how futures are created.
If $V$ is used by $C$ as the input of a future range, then the edge from
$V$ to $C$ represents what Ferrante {\em et al.} call a {\em control
dependence}~\cite[p.323]{Ferrante87}.
We use dashed lines to represent these edges.
All the other edges denote {\em data dependences}~\cite[p.322]{Ferrante87}.
As we will show later, control dependence edges increase the precision of our
algorithm to solve future bounds.

@@

\noindent
\textbf{Propagating Ranges in Topological Order. }
After building the constraint graph, we find its strongly connected components.
We collapse these components in super nodes, and then propagate
ranges along the resulting digraph.
This approach is essential for scalability, because all the complexity of our
algorithm lies in the resolution of strong components.
Our tests show that the vast majority of the strongly connected components are
singletons.
For instance, 99.11\% of the SCCs in SPEC CPU 2006 dealII (\texttt{447.dealII}) have
only one node.
Moreover, the composite components usually contain a small number of nodes.
As an example, the largest component in dealII has 2,131 nodes,
even though dealII's constraint graph has over one million nodes.
This large SCC exists due to a long chain of mutually recursive function calls.

@@

\noindent
\textbf{Future resolution: }
The next phase of the algorithm to determine intervals inside a strong
component consists in replacing futures by actual bounds, a task that we
accomplish by using the rules below:
%
\begin{eqnarray*}
\begin{array}{c}
\inferrule{Y = X \sqcap [l, \fun{ft}(V) + c] \\ \ub{I[V]} = u}
{Y = X \sqcap [l, u + c]} \mbox{\hspace{0.3cm}} u, c \in \mathbb{Z} \cup \{-\infty, \infty\}
\\
\\
\inferrule{Y = X \sqcap [\fun{ft}(V) + c, u] \\ \lb{I[V]} = l}
{Y = X \sqcap [l + c, u]} \mbox{\hspace{0.3cm}} l, c \in \mathbb{Z} \cup \{-\infty, \infty\}
\end{array}
\end{eqnarray*}
%
To correctly replace a future $\fun{ft}(V)$ that limits a constraint
variable $V'$, we need to have already applied the growth analysis onto $V$.
Had we considered only data dependence edges, then it would be possible
that $V'$'s strong component would be analyzed before $V$'s.
However, because of control dependence edges, this case cannot happen.
The control dependence edges ensure that any topological ordering of the
constraint graph either places $V$ before $V'$, or places these nodes
in the same strongly connected component.
For instance, in Figure~\ref{fig:ex_graph}, variables $j_1$ and $i_t$ are in
the same SCC only because of the control dependence edges.
Figure~\ref{fig:ex_partition_grow_crop}(c) shows the result of resolving
futures in our running example.
The information that we acquire from the growth analysis is essential in this
phase.
For instance, the growth analysis has found out that the value stored in variable
\texttt{i}$_i$ can only increase.
Given that this variable is assigned the initial value zero, we can replace
$\fun{ft}(I_1)$ with this value.

\noindent
\textbf{Narrowing: } the last step that we apply on the strongly connected
component is the narrowing phase.
In this step we use values extracted from conditional tests to restrict the
bounds of the constraint variables.
We use the narrowing operator firstly proposed by Cousot and
Cousot~\cite{Cousot77}, which we show below:
%
\begin{equation*}
I[Y] =
\begin{cases}
\mbox{if} \ \lb{I[Y]} = -\infty  \ \mbox{and} \ \lb{e(Y)} > -\infty \ \mbox{then} \ [\lb{e(Y)}, \ub{I[Y]}] \\
\mbox{elif} \ \ub{I[Y]} = \infty \ \mbox{and} \ \ub{e(Y)} < \infty \ \mbox{then}
\ [\lb{I[Y]}, \ub{e(Y)}] \\
\mbox{elif} \ \lb{I[Y]} > \lb{e(Y)} \ \mbox{then} \ [\lb{e(Y)}, \ub{I[Y]}] \\
\mbox{elif} \ \ub{I[Y]} < \ub{e(Y)} \ \mbox{then} \ [\lb{I[Y]}, \ub{e(Y)}]
\end{cases}
\end{equation*}
%
Figure~\ref{fig:ex_partition_grow_crop}(d) show the result of our narrowing
operator in our running example.
Ranges improve due to the two conditional tests in the program.
Firstly, we have that $I[I_t] = I[I_1] \sqcap [-\infty, 98]$, which gives us
that $I[I_t] = [0, 98]$.
We also have that $I[J_t] = I[J_1] \sqcap [0, \infty]$, giving
$I[J_t] = [0, 99]$.
From these new intervals, we can narrow the ranges bound to the other constraint
variables.

@@

Figure~\ref{fig:ex_ir} compares the u-SSA form with the SSA and e-SSA
intermediate program representations.
We use the notation $v = \bullet$ to denote a definition of variable $v$, and
$\bullet = v$ to denote a use of it.
Figure~\ref{fig:ex_ir}(b) shows the example program converted to the SSA format.
Different definitions of variable $u$ have been renamed, and a
$\phi$-function joins these definitions into a single name.
The SSA form sparsifies a dataflow analysis that only extracts information from
the definition sites of variables, such as constant propagation.
Figure~\ref{fig:ex_ir}(c) shows the same program in e-SSA form.
This time we have renamed variable $v$ right after the conditional test where
this variable is used.
The e-SSA form serves dataflow analyses that acquire information from definition
sites and conditional tests.
Examples of these analyses include array bounds checking
elimination~\cite{Bodik00} and traditional implementations of range
analyses~\cite{Gough94,Patterson95}.
Finally, Figure~\ref{fig:ex_ir}(d) shows our example in u-SSA form.
The live range of variable $v_1$ has been divided right after its use.
This representation assists analyses that learn information from the way
that variables are used, and propagate this information forwardly.

@@

%SUB instrumentation
The instrumentation that we insert is mostly straightforward.
We discuss in the rest of this section a few interesting cases.
When dealing with an unsigned \textsc{Sub} instruction, e.g,
$x = o_1 \ -_u \ o_2$, then a single check is enough to detect the bad
behavior: $o_1 < o_2$.
If $o_2$ is greater than $o_1$, then we assume that it is
a bug to try to represent a negative number in unsigned arithmetics.
Regarding multiplication, e.g., $x = o_1 \times o_2$, if $o_1 = 0$, then this
operation can never cause an overflow.
This test is necessary, because we check integer overflows in
multiplication via the inverse operation, e.g., integer division.
Thus, the test prevents a division by zero from happening.
The \textsc{Trunc} instruction, e.g., $x = \ \downarrow_n \ o_1$ assigns to $x$
the $n$ least significant bits of $o_1$.
The dynamic check, in this case, consists in expanding $x$ to
the datatype of $o_1$ and comparing the expanded value with $o_1$.
The LLVM IR provides instructions to perform these type expansions.
Note that our instrumentation catches any truncation that might result in
data loss, even if this loss is benign.
To make the dynamic checks more liberal, we give users the possibility of
disabling tests over truncations.

\paragraph{Practical Considerations: }
%Overview about the instrumentation technique
Our instrumentation library inserts new instructions into
the target program.
Although the dynamic check depends on the instruction that is instrumented,
the general modus operandi is the same.
Dynamic tests check for overflows after they happen.
The code that we insert to detect the overflow diverts the program flow in case
such an event takes place.
Figure~\ref{fig:instrumented_cfg} shows an actual control flow graph, before
and after the instrumentation.
Clearly the instrumented program will be larger than the original code.
Figure~\ref{fig:num_instructions} shows how many LLVM instructions are
necessary to instrument each arithmetic operation.
These numbers do not include the instructions necessary to handle the overflow
itself, e.g., block \texttt{\%11} in Figure~\ref{fig:instrumented_cfg}, as this
code is not in the program's main path.
Nevertheless, as we show empirically, this growth is small when
compared to the total size of our benchmarks, because most of the instructions
in these programs do not demand instrumentation.
Furthermore, none of the instructions used to instrument integer arithmetics
access memory.
Therefore, the overall slowdown that the instrumentation causes is usually
small, and the experiments in Section~\ref{sub:inst} confirm this
observation.

@@

Which actions are performed once the overflow is detected depends on the
user, who has the option to overwrite the \texttt{handle\_ overflow}
function in Figure~\ref{fig:instrumented_cfg}.
Our library gives the user three options to handle overflows. 
First option: no-op.
This option allows us to verify the slowdown produced by the new instructions.
Second option: logging.
This is the standard option, and it preserves the behavior of the instrumented
program.
Whenever an overflow is detected, we print \texttt{Overflow detected in
FileName.cpp, line X.} in the standard error stream.
Third option: abort.
This option terminates the program once an overflow is detected.
Thus, it disallows undefined behavior due to integer overflows, and gives us the
opportunity to use the u-SSA form to get extra precision.

@@

\noindent
\textbf{Time and Memory Complexity: }
Figure~\ref{fig:TimeCorr} compares the time to run our range analysis with the size of
the input programs.
We show data for the 100 largest benchmarks in our test suite, considering the number
of variable nodes in the constraint graph.
We perform function inlining before running our analysis.
Each point in the X line corresponds to a benchmark.
We analyze the smallest benchmark in this set, \texttt{Prolangs-C/deriv2}, which
has 1,131 variable nodes in the constraint graph, in 20ms.
We take 9.91 sec to analyze our largest benchmark, \texttt{403.gcc}, which,
after function inlining, has 1,419,456 assembly instructions, and a
constraint graph with 679,652 variable nodes.
For this data set, the coefficient of determination $(R^2)$ is 0.967, which
provides very strong evidence about the linear asymptotic complexity of our
implementation.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/TimeCorr}
\end{center}
\caption{\label{fig:TimeCorr}
Correlation between program size (number of var nodes in constraint
graphs) and analysis runtime (ms).
Each point represents a benchmark.
Coefficient of determination = 0.967.
}
\end{figure}

The experiments also reveal that the memory consumption of our implementation
grows linearly with the program size.
Figure~\ref{fig:MemCorr} plots these two quantities.
The linear correlation, in this case, is even stronger than that found in
Figure~\ref{fig:TimeCorr}: the coefficient of determination is 0.9947.
The figure only shows our 100 largest benchmarks.
Again, SPEC \texttt{403.gcc} is the largest benchmark, requiring
265,588KB to run.
Memory includes stack, heap and the executable program code.

@@

\noindent
\textbf{Precision: }
Our implementation of range analysis offers a good tradeoff between
precision and runtime.
Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance,
takes about 25 minutes to go over a program with almost 900 basic blocks.
We analyze programs of similar size in less than one second.
We do not claim our approach is as precise as such algorithms, even though we
are able to find exact bounds to 4/5 of the examples presented
in~\cite{Lakhdar11}.
On the contrary, we present a compromise between precision and speed
that scales to very large programs.
Nevertheless, our results are not trivial.
We have implemented a dynamic profiler that measures, for each variable,
its upper and lower limits, given an execution of the target program.
Figure~\ref{fig:precision} compares our results with those measured
dynamically for the Stanford benchmark, which is publicly
available in the LLVM test suite.
We chose Stanford because these benchmarks do not read data from external
files; hence, imprecisions are due exclusively to library functions that we cannot
analyze.

@@

We have classified the bounds estimated by the static analysis into four
categories.
The first category, called $1$, contains tight bounds:
during program execution, the variable has been assigned an upper,
or lower limit, that equals the limit inferred statically.
The second category, called $n$, contains the bounds that are
within twice the value inferred statically.
For instance, if the range analysis estimates that a variable $v$ is in the
range $[0, 100]$, and during the execution the dynamic profiler finds that
its maximum value is $51$, then $v$ falls into this category.
The third category, $n^2$, contains variables whose actual value is within
a quadratic factor of the estimated value.
In our example, $v$'s upper bound would have to be at most $10$ for it to
be in this category.
Finally, the fourth category contains variables whose estimated value lays
outside a quadratic factor of the actual value.
We call this category {\em imprecise}, and it contains mostly the limits that
our static analysis has marked as either $+\infty$ or $-\infty$.
As we see in Figure~\ref{fig:precision}, 54.11\% of the lower limits that
we have estimated statically are exact.
Similarly, 51.99\% of our upper bounds are also tight.
The figure also shows that, on average, 37.39\% of our lower limits are
imprecise, and 35.40\% of our upper limits are imprecise.
This result is on pair with those obtained by more costly analysis, such as
Stephenson {\em et al.}'s~\cite{Stephenson00}.
